Mergeable Types

### 2017-02-28 paper/implementation.tex
edits

\name programming model is realized on top of Irmin~\cite{irmin}, an OCaml
library database implementation that is part of the MirageOS
project~\cite{mirage}. Irmin provides a persistent multi-versioned store with
with a content-addressable heap abstraction. Simply put, content-addressability
means that the address of a data block is determined by its content. If the
content changes, then so does the address. Old content continues to be
available the old address. Content-addressability also results in constant time
structural equality checks, which we exploit in our mergeable rope
implementation (Section~\ref{sec:ropes}).

Irmin provides support for distribution, fault-tolerance and concurrency
control by incorporating Git distributed version control~\cite{git} protocol
over its object model. Indeed, Irmin is fully compatible with the Git command
line tools. Distributed replicas in \name are created by cloning a \name
repository. Due to \name's mergeable type, each replica can operate completely
independently, accepting client requests, even when disconnected from other
replicas, resulting in a fully available distributed database system.

@@

Concurrent operations in Irmin are tracked within the notion of branches,
allowing the programmer to explicitly merge the branches on demand. \name
realizes the concurrent programming model on top of the notion of branches;
each replica operates on its own branch, and thus is isolated from actions on
other replicas. In presence of network partitions, the nodes in one partition
may not receive updates from the other partition, hence merges cannot happen
across a partition. This can be a problem if a pair of branches in one
partition (the current partition) have multiple LCAs in the other partition
(the remote partition), hence cannot merge. Fig.~\ref{fig:partitions}
illustrate this problem for the branching structure of
Fig.~\ref{fig:external-lcas}.

Fortunately, the access to full (causal) branching history assured by Irmin at
every node comes to the rescue. Relying on the history, the current partition
can fork-off (\C{fork}) new branches that start where the remote branches left,
and map them to new (virtual) nodes that emulate the nodes in the remote
partition. For the example in Fig.~\ref{fig:partitions}, the partition of $b_1$
and $b_4$ can fork new branches $b_2'$ and $b_3'$ from $b_2$ and $b_3$,
respectively, and resume the activity as if the partition never happened. After
merging the first versions on $b_2'$ and $b_3'$, branches $b_1$ and $b_2$ will
be in the same state as before (when the partition happened), except that the
branches are now mergeable because $b_2'$ and $b_3'$ can merge. The ability to
track full provenance information is thus crucial for \name to overcome network
partitions, making it an appropriate programming model for highly-available
replicated data types. \name monad conceals the branching structure, but also
transparently performs the necessary merges to obtain a history graph with a
unique lowest common ancestor.

Importantly, Irmin supports user-defined three-way merge functions for
reconciling concurrent operations. While Irmin's merge functions are defined
over objects on Irmin's content-addressable heap, \name's merge functions are
defined over OCaml types. We address this representational mis-match with the
help of OCaml's PPX metaprogramming support~\cite{ppx} to derive bi-directional
transformations between objects on OCaml and Irmin heaps. We also derive the
various serialization functions required Irmin. This address the representation
mismatch between the application and storage layers. Synchronization between
replicas is performed using Git's notion of pushing and pulling updates from
remotes over the Git transfer protocol~\cite{git-tp}. As a result, we reap the
benefits of efficient delta-transfer (only missing objects are transferred
between replicas), compression and end-to-end encrypted communication between
the replicas.

>>>>

The \name programming model is realized on top of Irmin~\cite{irmin},
an OCaml library database implementation that is part of the MirageOS
project~\cite{mirage}. Irmin provides a persistent multi-versioned
store with a content-addressable heap abstraction. Simply put,
content-addressability means that the address of a data block is
determined by its content. If the content changes, then so does the
address. Old content continues to be available from the old
address. Content-addressability also results in constant time
structural equality checks, which we exploit in our mergeable rope
implementation (Section~\ref{sec:ropes}), among others.

Irmin provides support for distribution, fault-tolerance and
concurrency control by incorporating the Git distributed version
control~\cite{git} protocol over its object model. Indeed, Irmin is
fully compatible with Git command line tools. Distributed replicas in
\name are created by cloning a \name repository. Due to \name's
support for mergeable types, each replica can operate completely
independently, accepting client requests, even when disconnected from
other replicas, resulting in a highly available distributed system.

@@

Concurrent operations in Irmin are tracked using the notion of
branches, allowing the programmer to explicitly merge branches on
demand. \name's concurrency and distribution model is realized in
terms of branches; each replica operates on its own branch, and thus
is isolated from the actions on other replicas. In the presence of
network partitions, the nodes in one partition may not receive updates
from the other partition, preventing merges from happening across a
partition. This can be a problem if a pair of branches in one
partition (the current partition) have multiple LCAs in the other
partition (the remote partition), and hence cannot
merge. Fig.~\ref{fig:partitions} illustrate this problem for the
branching structure of Fig.~\ref{fig:external-lcas}.

Fortunately, access to full (causal) branching history assured by
Irmin at every node comes to the rescue. Relying on the history, the
current partition can fork-off (\C{fork}) new branches that start
where the remote branches left, and map them to new (virtual) nodes
that emulate the nodes in the remote partition. For the example in
Fig.~\ref{fig:partitions}, the partition of $b_1$ and $b_4$ can fork
new branches $b_2'$ and $b_3'$ from $b_2$ and $b_3$, respectively, and
resume the activity as if the partition never happened. After merging
the first versions on $b_2'$ and $b_3'$, branches $b_1$ and $b_2$ will
be in the same state as before (when the partition happened), except
that the branches are now mergeable because $b_2'$ and $b_3'$ can
merge. The ability to track full provenance information is thus
crucial for \name to overcome network partitions, making it an
appropriate programming model for highly-available replicated data
types. The \nameMonad conceals branching structure, but also
transparently performs the necessary merges to obtain a history graph
with a unique lowest common ancestor.

Importantly, Irmin supports user-defined three-way merge functions for
reconciling concurrent operations. While Irmin's merge functions are defined
over objects on Irmin's content-addressable heap, \name's merge functions are
defined over OCaml types. We address this representational mis-match with the
help of OCaml's PPX metaprogramming support~\cite{ppx} to derive bi-directional
transformations between objects on OCaml and Irmin heaps. We also derive the
various serialization functions required by Irmin. This address the representation
mismatch between the application and storage layers. Synchronization between
replicas is performed using Git's notion of pushing and pulling updates from
remotes over the Git transfer protocol~\cite{git-tp}. As a result, we reap the
benefits of efficient delta-transfer (only missing objects are transferred
between replicas), compression and end-to-end encrypted communication between
the replicas.
